Low Infant Birth Weight in Brazil

### 2024-02-23 code/processing-code/processingfile-v2.qmd
Part 2
Finished cleaning and EDA for Part 2 of project.

---
title: "Cleaning and Processing"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

Please note I used ChatGPT and CoPilot throughout to help troubleshoot code.

@@

This contains the same code and comments/information as `processingcode.R`. This quarto file does not pull code from the R script but rather includes it directly.

@@

Load needed packages. Make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
```

@@

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","MaternidadeLaranjeiras.csv")
ML_raw <- read_csv(data_location)

# Structure of Data
glimpse(ML_raw)
```
From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.

@@

First we can look at the Codebook and practice table making.
```{r}
# Path to codebook
codebook_location <- here::here("data", "raw-data", "Codebook.csv")
codebook <- read_csv(codebook_location)

# Full column specification
spec(codebook)

# Display the structure of the data frame
str(codebook)

# Display the first few rows of the data frame
head(codebook)

# Create table
codebook_table <- kable(codebook, caption = "Codebook") %>% kable_styling()

#Print table
print(codebook_table)
```

@@

```{r}
# Looking at the data
dplyr::glimpse(ML_raw)
summary(ML_raw)
head(ML_raw)
skimr::skim(ML_raw)
```

@@

```{r}
# Remove unnecessary variables
ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Birth_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
```

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
```
We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations. This is a lot. Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.

```{r}
# Remove variables with missing data not relevant to hypothesis
ML_2 <- ML_1 %>% dplyr::select(-CivilStatus, -GestationalAge_Months)
```

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_2 %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(missing_data_summary)
```

@@

```{r}
# Transform character variables into factors
ML_3 <- ML_2 %>%
  mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex), as.factor))

# Check variables
glimpse(ML_3)

# Check levels of factor variables
map(ML_3[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')], levels)

# Define factor variables to check frequencies
factor_variables <- c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')

# Check frequencies of factor variables
frequency_table <- ML_3 %>%
  count(across(all_of(factor_variables)))

print(frequency_table)
```

@@

```{r} 
# Create dataset for summary statistics (includes missing values)
ML_summary <- ML_3

# Creating dataset for linear regression
ML_linear <- ML_3 %>% tidyr::drop_na()

# Check for no missing data in linear model
ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(ML_linear_summary)
```
Now, I can see that my `ML_linear` dataset has no missing data. It has 2032 observations and 11 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.

@@

Finally, I will save the clean data both the `ML_Linear` dataset and the `ML_summary` dataset as RDS files
.
```{r}
# Save ML_Linear as RDS
save_data_location <- here::here("data","processed-data","ML_linear_processed.rds")
saveRDS(ML_linear, file = save_data_location)

# Save ML_summary as RDS
save_data_location <- here::here("data","processed-data","ML_summary_processed.rds")
saveRDS(ML_summary, file = save_data_location)
```

>>>>

---
title: "Cleaning and Processing with code pulled in"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

Please note I used ChatGPT and CoPilot throughout to help troubleshoot code. I'm not sure that I did this file and the `processingcode.R` files correctly. `processingfile-v1.qmd` is correct.

@@

This contains the same code and comments/information as `processingcode.R`. This quarto file pulls in from the R script using the code chunk labels.

@@

This needs to run to load the R script so we can include the code chunks below

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('processingcode.R')
```

```{r, packages, message = FALSE, warning = FALSE}
```

@@

```{r, loaddata}
```
From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.

@@

First we can look at the Codebook and practice table making.
```{r, checkdata}
```

@@

```{r, exploredata}
```

@@

```{r, cleandata}
```

@@

We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations. This is a lot. Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.

@@

@@

@@

```{r, createdatasets}
```
Now, I can see that my `ML_linear` dataset has no missing data. It has 2032 observations and 11 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.

@@

Finally, I will save the clean data both the `ML_Linear` dataset and the `ML_summary` dataset as RDS files
.
```{r, savedata}
```

### 2024-02-23 products/manuscript/Manuscript_Cassia.qmd
Part 2
Finished cleaning and EDA for Part 2 of project.

---
title: "Low Infant Birth Weight in Brazil"
subtitle: "A Historical Data Analysis Approach"
author: Cassia Roth
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
    bibliography: assets/Mada_Bib.bib
    csl: assets/vancouver.csl
---

```{r, echo=FALSE, message=FALSE}
#Loading R packages
library(here)
library(knitr)
library(renv)
library(tidyverse)
```

# Summary/Abstract
This project uses a linear regression model to understand disparities in infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s. I will use a linear model to estimate the relationship of maternal variables on infant birth weight. I hypothesize that infants born to women of color will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.

@@

## General Background Information
Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality. Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts. This is a significant public health concern because low birth weight is associated with higher infant mortality, and with long-term health problems such as diabetes, hypertension, and heart disease.

@@

## Description of data and data source
I will analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro. The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*. In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*. In 1922, it became the *Revista de Gynecologia e D’Obstetricia*.

The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). From the available issues, I recorded the following information, when available, for all patients: patient number, gravidity and parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history. Gravidity refers to the total number of pregnancies a woman has, regardless of duration. Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posnerOxornFooteHumanLabor2013].

Clinical notes only sporadically included infant demographic information for spontaneous abortions or stillbirths. For spontaneous abortions, stillbirths, and live births, I recorded, when available, infant sex, weight, length. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy. However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraudMauriceauManeuverDeleterious1997), coded as operatory or natural by physicians, as interventionist.

@@

## Questions/Hypotheses to be addressed
I will quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, Maternidade Laranjeiras, in the 1920s. I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.

The outcome I will measure is infant birth weight. The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganizationInternationalClassificationDiseases2022]. I will use the WHO’s classification to categorize birth weight.

@@

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
```{r}
#Setting working directory using here package
here("data/raw-data/MaternidadeLaranjeiras.csv")

#Reading in file
ML <- read.csv(here("data/raw-data/MaternidadeLaranjeiras.csv"))

#Looking at structure of data
str(ML)
```
/Users/cassiaroth/Documents/GitHub/MADARoth/Roth-MADA-project/MaternidadeLaranjeiras.csv

```{r}
```
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

@@

# Results

@@

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

@@

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@@

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

@@

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

>>>>

---
title: "Low Infant Birth Weight in Brazil"
subtitle: "A Historical Data Analysis Approach"
author: Cassia Roth
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
    bibliography: ../../assets/Mada_Bib.bib
    csl: ../../assets/vancouver.csl
---

```{r, echo=FALSE, message=FALSE}
#Loading packages
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
```

# Summary/Abstract
This cross-sectional study uses a linear regression model to understand disparities in infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s. I will use a linear model to estimate the relationship of maternal variables on infant birth weight. I hypothesize that infants born to women of color will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.

@@

## General Background Information
Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality [@travassos2004]. Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts. This is a significant public health concern because low birth weight is associated with higher infant mortality, and with long-term health problems such as diabetes, hypertension, and heart disease.

@@

## Description of data and data source

I will analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro. The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*. In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*. In 1922, it became the *Revista de Gynecologia e D’Obstetricia*.

The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). From the available issues, I recorded the following information, when available, for all patients: patient number, gravidity and parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history. Gravidity refers to the total number of pregnancies a woman has, regardless of duration. Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013].

Clinical notes only sporadically included infant demographic information for spontaneous abortions or stillbirths. For spontaneous abortions, stillbirths, and live births, I recorded, when available, infant sex, weight, length. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy. However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraud1997), coded as operatory or natural by physicians, as interventionist.

@@

## Questions/Hypotheses to be addressed

I will quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, Maternidade Laranjeiras, in the 1920s. I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.

The outcome I will measure is infant birth weight. The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022]. I will use the WHO’s classification to categorize birth weight.

@@

## Data import and cleaning

Please see the qmd.file `processingfile-v1.qmd1` for full detail on the data import and cleaning process.

@@

@@

The exploratory analysis was performed on the ML_summary dataset with n = 2845 observations and 14 variables. Please see the Codebook for more details on variables.

There were three continuous variables: maternal age, infant birth weight, and infant length. I am aware that this table is not rendering correctly in the word document, but I am unable to figure out why. If I load library(gt), then my word document is corrupted and will not open at all. I will continue to work on this issue. 
```{r}
#| label: tbl-summarytablecont
#| tbl-cap: "Continous Data Summary Table"
#| echo: FALSE
summary_stats_table <- readRDS(here::here("results", "tables", "summary_stats_table.rds"))
summary_stats_table
```

Birth weight is the outcome variable of interest, so let's look at some visualizations of the data. Here is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO defines as normal birth weight.

```{r}
#| label: fig-schematic-1
#| fig-cap: Histogram of birth weight
#| echo: FALSE
knitr::include_graphics(here("results","figures","weight_distribution.png"))
```

Birth length is less important for our analysis, but we might want to look at the relationship between weight and length. Below we can see a histogram of the distribution of birth lengths in the sample. The red line marks the average birth length for both male and female infants (49 centimeters).

```{r}
#| label: fig-schematic-2
#| fig-cap: Histogram of birth length
#| echo: FALSE
knitr::include_graphics(here("results","figures","length_distribution.png"))
```

Finally, we will look at the relationship between infant weight and length, both not-stratified and stratified by sex. Unsurprisingly, birth weight and length are positively correlated.

```{r}
#| label: fig-schematic-3
#| fig-cap: Scatterplot of birth weight by length
#| echo: FALSE
knitr::include_graphics(here("results","figures","length_weight.png"))
```
Finally, since male infants are usually slightly heavier than females, let's look at a scatterplot of birth weight by length, stratified by sex.

```{r}
#| label: fig-schematic-4
#| fig-cap: Scatterplot of birth weight by length stratified by sex
#| echo: FALSE
knitr::include_graphics(here("results","figures","length_weight_sex.png"))
```

For information on the categorical variables, please see eda-v1.qmd. Please note Dr. Handel, I am really struggling with making a table of these variables, and I will continue working on that moving forward. If you are interested in the code and data, please see the eda-v1.qmd file.

@@

@@

@@

## Summary and Interpretation

## Strengths and Limitations

## Conclusions
